Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit

### 2016-07-01 paper/paper.tex
Misc text additions and corrections

\documentclass[10pt,twocolumn]{article}
\usepackage{graphicx}
\usepackage{fancyhdr}
\usepackage{amssymb,amsmath}
\usepackage[usenames,dvipsnames]{color}
\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
\usepackage{doi}
\usepackage{setspace}
\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, right=1in]{geometry}
\usepackage[compact]{titlesec}
\usepackage{abstract}
\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}
\usepackage[authoryear,square]{natbib}
\usepackage{aas_macros}
\usepackage{array}
\pagestyle{fancy}
\fancyhead[R]{Veibell \thepage}
\fancyhead[L]{}

\newcolumntype{L}{>{$}l<{$}} %To enable math mode in tables (for saying things like B_z)
\newcolumntype{C}{>{$}c<{$}}
\newcommand{\vinote}[1]{\textcolor{red}{\textbf{#1}}} %If you want \inote{#} instead of just \note #, for inline notes
\def\vnote#1\par{\textcolor{red}{\textbf{#1}}\\} %Make note bold and red, re-add newline
\newcommand{\req}{\ensuremath{\rho_{eq}}}
\newcommand{\inote}[1]{\textcolor{blue}{\textbf{#1}}} %Inline notes in blue
\def\note#1\par{\textcolor{blue}{\textbf{#1}}\\}

\begin{document}
\title{Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit}
\author{
Victoir Veibell\footnote{vveibell@gmu.edu}
\and
R.S. Weigel\footnote{rweigel@gmu.edu}}

\setlength{\parskip}{3ex}
\renewcommand{\labelitemi}{$-$}
\titlespacing{\section}{0pc}{0.2pc}{-1pc}
\titlespacing{\subsection}{0pc}{0.1pc}{-1pc}
\titleformat*{\section}{\normalsize\bfseries}
\titleformat*{\subsection}{\small\bfseries}

\twocolumn[
  \begin{@twocolumnfalse}
\maketitle
\hrule
\begin{abstract}
In this work we consider two types of events, determined by decreases in $D_{st}$ below a threshold value or increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is no statistically significant.  Although very large geomagnetic storms have been observed to correspond to large increases in the $\rho_{eq}$, in general $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or the solar wind parameters $B_{z}$ and $P_{dyn}$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.
\end{abstract}
\vspace{2em}
\hrule
\vspace{2em}
 \end{@twocolumnfalse}
  ]

\saythanks

@@

The plasma trough is a transition region between the higher density and lower temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature}.  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapause, the density has contributions from the co-rotating plasmasphere and Earthward convecting magnetosphere plasmas.

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966}.  There are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}; \cite{LemaireEarthsPlasmasphere}).

The dependence of the plasma trough density on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of hydrogen ion density found a significant dependence on the plasmasphere boundary on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but there the relationship between geomagnetic activity and density in the plasma trough was not explicitly considered.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 (observations were made when CRRESS was in an elliptical, near equatorial orbit and primarily in the plasma trough region). They found that the average ion mass, $M=\rho/n_e$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of 12:00 and 18:00 with most observations having $L$ between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of $\rho$ using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for the bins of $D_{st}$ (-142 to -31 nT and -31 to 37 nT) and $K_p$ (1.5 to 3.4 and 3.4 to 5.9) considered .

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density $\rho$ was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}L/(R/R_E)$.  They found a high correlation ($\sim 0.93$) between 27-day averages of F10.7 and 27-day medians of $\rho_{eq}$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, an enhancement in $\rho_{eq}$ appeared on the same day as minimum $D_{st}$.

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation in the plasma trough was not calculated, the data presented for four events show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

\cite{Denton2016} ...

The results in the papers discussed indicate that (1) the mass density in the plasma trough region is best correlated with $F10.7$ and (2) there exists a much weaker relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Takahashi2010} and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density in the plasma trough region.  In addition, we attempt to identify solar wind processes that may drive changes in mass density.

There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, and geomagnetic and solar wind parameters also correlate with F10.7, (3) the dependence of mass density on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, corresponding to low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of mass density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in mass density associated with strong southward IMF and enhanced magnetospheric activity (\cite{Takahashi2010}; \cite{Yao2008}).

Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 dependence.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northward and southward IMF.

In addition, we consider the influence of time averaging by comparing results using daily means and medians and hourly means and medians of solar wind and magnetospheric parameters.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ are from the dataset of \cite{Denton}, which was used in \cite{Takahashi2010}; data are available from 1980 through 1991 from GOES 2, 3, 5, 6 and 7.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly median over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used.  Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot}.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites are 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmasphere earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in the analysis in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events.  Error bars for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values used.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 329 events from 1989-1991.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  For the non-\req\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of \req\ values as we used all available measurements instead of restricting to only values where a non--fill $\rho_{eq}$ value existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6, which spans 1983-1991 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.  
To determine if the magnitude of the density medians in the top panel of \ref{DailyAveragedDstEvents} on days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each day was created by sampling the actual values with replacement.  The observed difference between the medians is shown in the table along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.  The results are shown in the fourth column of Table~\ref{BootstrapDifferenceTable}. The fourth column shows the results of a 2-population t-test and the fifth column indicates ???.

If a statistically significant difference in the medians between two days existed, the fraction is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the \% value shown in the table; the highest confidence level of 87.6\% is in the difference between day 0 and day 1.

@@

%\input{../tables/DeltaBootstraps.txt}

\begin{table}
\small

\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

We have also considered hourly averages as shown in the top panel of Fig.~\ref{HourlyAveragedDstEvents}.  A total of 668 events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.  The error bars were computed in the same way as Fig.~\ref{DailyAveragedDstEvents}.  In this plot, the overall trend is similar to that found in the bottom panel of Fig.~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small relative to the error bars around the time of the event onset.  At this timescale no obvious change in mass density at hourly timescale is visible.

The bottom panel of Fig.~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there were an average of \inote{X} samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}.  To determine of this relationship can be observed in the events of the top panel of Fig.~\ref{HourlyAveragedDstEvents}, we sorted the events by the average $D_{st}$ values in the time interval of 0-12 hours, split the events in two and recomputed the epoch averages of $\rho_{eq}$.  The result is that there is no statistically significant difference in the mass density epoch averages for large versus small events.  This indicates that to confirm a statitical $\rho_{eq}$ dependence on large geomagnetic storms would require a dataset with a much larger number of strong geomagnetic storms.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho\ $.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\req\ $ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, all fill values were replaced with linearly interpolated values. \vinote{Not including \req\ events?}

Here we see that such events tend to be associated with positive $B_z^{IMF}$ and low geomagnetic activity (\cite{Denton2016}).

To determine if the sign of $B_z^{IMF}$ influences the $\rho\ $ epoch time series, we separated the events by the average of $B_z^{IMF}$ in a time window of $4$~hours.

Figure~\label{fig:RhoBinned}(a) shows the epoch curves for $B_z^{IMF}\geq=0$ (257) events and $B_z^{IMF}\ge=0$ (139) have no statistically significant difference.

>>>>

%\documentclass[11pt,twocolumn]{article}
%\usepackage{graphicx}
%\usepackage{fancyhdr}
%\usepackage{amssymb,amsmath}
%\usepackage[usenames,dvipsnames]{color}
%\usepackage[colorlinks,citecolor=RedViolet,urlcolor=blue]{hyperref}
%\usepackage{doi}
%\usepackage{setspace}
%\usepackage[paperwidth=8.5in, paperheight=11in,top=1in, bottom=1in, left=1in, r%ight=1in]{geometry}
%\usepackage[compact]{titlesec}
%\usepackage{abstract}
%\usepackage[position=t,singlelinecheck=off,justification=raggedleft]{subcaption}
%\usepackage[authoryear,square]{natbib}
%\usepackage{aas_macros}
%\usepackage{array}
%\pagestyle{fancy}
%\fancyhead[R]{Veibell \thepage}
%\fancyhead[L]{}

%\newcolumntype{L}{>{$}l<{$}} %To enable math mode in tables (for saying things like B_z)
%\newcolumntype{C}{>{$}c<{$}}
%\newcommand{\vinote}[1]{\textcolor{red}{\textbf{#1}}} %If you want \inote{#} instead of just \note #, for inline notes
%\def\vnote#1\par{\textcolor{red}{\textbf{#1}}\\} %Make note bold and red, re-add newline
%\newcommand{\req}{\ensuremath{\rho_{eq}}}
%\newcommand{\inote}[1]{\textcolor{blue}{\textbf{#1}}} %Inline notes in blue
%\def\note#1\par{\textcolor{blue}{\textbf{#1}}\\}

\documentclass[12pt]{article}
\bibliographystyle{./style/agufull}
\usepackage[square,comma,authoryear]{natbib}
\usepackage{epsfig}
\usepackage{style/lineno}
\usepackage{amsmath}
%\usepackage{booktabs}
\linenumbers*[1]
\usepackage{subfig}

\usepackage{setspace}  
\usepackage{fancyhdr}

\newcommand{\citeay}[1]{%
\citeauthor{#1}, \citeyear{#1}%
}

\pagestyle{fancy}
\fancyhead{}
\fancyfoot{}
\fancyhead[R]{X - \thepage}
\fancyhead[L]{VEIBELL: MASS DENSITY AT GEO}
\fancyfoot[L]{D R A F T}
\fancyfoot[C]{\today}
\fancyfoot[R]{D R A F T}
\renewcommand{\headrulewidth}{0.0pt}
\renewcommand{\footrulewidth}{0.0pt}

\oddsidemargin 0.0in
\evensidemargin 1.0in
\textwidth 6.5in
\headheight 0.1in
\topmargin -0.5in
\textheight 9.0in

\begin{document}
\title{Solar wind and geomagnetic activity influence on the equatorial plasma mass density at geosynchronous orbit}
\author{V. Veibell and R.S. Weigel}
\doublespacing
\maketitle

%\author{
%Victoir Veibell\footnote{vveibell@gmu.edu}
%\and
%R.S. Weigel\footnote{rweigel@gmu.edu}}

%\setlength{\parskip}{3ex}
%\renewcommand{\labelitemi}{$-$}
%\titlespacing{\section}{0pc}{0.2pc}{-1pc}
%\titlespacing{\subsection}{0pc}{0.1pc}{-1pc}
%\titleformat*{\section}{\normalsize\bfseries}
%\titleformat*{\subsection}{\small\bfseries}

%\twocolumn[
%  \begin{@twocolumnfalse}
%\maketitle
%\hrule
\begin{abstract}
We consider two types of events, identified by decreases in $D_{st}$ below a threshold value and increases in the the equatorial mass density at geosynchronous altitudes, $\rho_{eq}$, above a threshold value on 1-hour and 1-day time scales using the \cite{Takahashi2010} dataset.  From the $D_{st}$ events, we find that there is a weak and small amplitude ($< 10$\%) difference between $\rho_{eq}$ on the day of the event and the days before and after on both daily and hourly timescales.  Although very large geomagnetic storms have been observed to correspond to large increases in $\rho_{eq}$, on average, $\rho_{eq}$ following the onset of a geomagnetic storm does not depend on the storm intensity or north-south component of the interplanetary magnetic field, $B_z$. From the $\rho_{eq}$ events, we find a weak dependence on $B_{z}$ after the onset of an event, with higher average $B_{z}$ six hours after the event onset corresponding to larger $\rho_{eq}$ events.  In contrast, the average of $B_z$ six hours prior to $\rho_{eq}$ events does not have a statistically significant impact on $\rho_{eq}$ after onset.
\end{abstract}
%\vspace{2em}
%\hrule
%\vspace{2em}
% \end{@twocolumnfalse}
%  ]

%\saythanks

@@

%The plasma trough is a transition region between the higher density and lower temperature plasmasphere and the lower density and higher energy magnetosphere \citep{Mayr1968ModelMagnetosphereTemperature} and is .  Inside of the plasmapause, the density is primarily due to ionospheric photoionization; outside of the plasmapause, the density has contributions from the co-rotating plasmasphere and Earthward convecting magnetosphere plasmas.

The dependence of the plasmapause location on geomagnetic activity was observed as early as \cite{Carpenter1966} and there are well--validated models of the plasmapause boundary, as a function of $L$, $MLT$, and geomagnetic activity (\cite{LemaireEarthsPlasmasphere}, \cite{Moldwin2002ModelPlasmapause}, \cite{OBrien2003EmpiricalPlasmapause}).  Models also exist for the plasmasphere density (\cite{Gallagher1988EmpiricalModelPlasmasphere}; \cite{LemaireEarthsPlasmasphere}) with similar parameterization.

The dependence of the plasma density outside of the nominal plasmapause location on geomagnetic and solar wind conditions has not been extensively studied until the past decade (\cite{Takahashi2006}; \cite{Takahashi2010}; \cite{Denton2016}). Earlier works that considered profiles of hydrogen ion density found a significant dependence of the plasmasphere and plasmapause on geomagnetic activity (\cite{Chappell1970}; \cite{Maynard1977}; \cite{Carpenter1992}), but the relationship between geomagnetic activity and density outside of the plasmapause was not explicitly considered.

\cite{Takahashi2006} estimated magnetospheric mass density using measurements from the CRRES satellite during a 73-day period in 1991 when CRRES was in an elliptical, near equatorial orbit and primarily beyond the plasmapause. They found that the local average ion mass, $M$, had some correlation with geomagnetic activity, with more negative hourly $D_{st}$ values corresponding to higher $M$ and 1.5- and 3-day averages of $K_p$ corresponding to higher $M$ (3-hour averages of $K_p$ had little visual correlation with $M$). The mass density estimates were obtained from CRRES observations in the MLT range of [12:00, 18:00] with most observations at $L$ between 5 and 7 $R_E$.

\cite{Denton2006} found a weak relationship between $D_{st}$ and and $K_p$ and the field line distribution of mass density, $\rho$, using observations of the first three toroidal Alfv\'en harmonic frequencies for $L$ in the range of 6-8~$R_E$.  A more pronounced peak in the distribution along a filed line near the equator was found for lower $D_{st}$ and larger $K_p$.  The equatorial value of $\rho$ appeared identical for $D_{st}$ bins of [-142, -31] and [-31, 37] nT and $K_p$ bins of [1.5,3.4] and [3.4, 5.9].

\cite{Takahashi2010} developed a mass density dataset using measurements from the Space Environment Monitor instruments on the Geostationary Operational Environmental Satellites (GOES) satellites from 1980 through 1992, with most measurements in the range of $L=6.8\pm0.2 R_E$. The mass density was estimated using the the Alfv\'en wave velocity relationship, $V_A=B/\sqrt{\mu_0\rho}$, a magnetic field model, and a numerical solution to a wave equation with an ionospheric boundary and the assumption of a zero resistance ionosphere.  The equatorial mass density, $\rho_{eq}$, was derived from the estimated mass density using a power law dependence on the geocentric distance to the field line of the observation, $R$, $\rho=\rho_{eq}(LR_E/R)^{1/2}$.  They found a high correlation ($\sim 0.94$) between 27-day averages of $F10.7$ and 27-day medians of $\rho_{eq}$.

\cite{Takahashi2010} noted that downward spikes in the $D_{st}$ index coincided with with significant changes in $\rho_{eq}$ for $L$ near 6.8~$R_E$. For five storms, two had $\rho_{eq}$ spikes after the $D_{st}$ drop, two had $\rho_{eq}$ spikes before the drop, and one showed little change in $\rho_{eq}$.  A key result was that when daily-averaged measurements were considered, the epoch average of $D_st$ exhibited for storms with a minimum $D_{st} < -50$~nT, an enhancement in $\rho_{eq}$ the same day as minimum $D_{st}$ epoch averages.

\cite{Yao2008} studied the relationship between $D_{st}$ and the number density of low-energy $O^+$ in different regions (ring current and plasma sheet) using the TEAMS and ESA instruments on the low-altitude and high-inclination orbiting FAST satellite and found that the average $N_{O^+}$ across the sampled L-shells (2-14) had a strong correlation (0.88) with the minimum $D_{st}$ of the storm.  Although the correlation at geosynchronous distances was not calculated, the data presented for four events show that the variations in $N_{O^+}$ tended to appear near $D_{st}$ minimum ($\pm$ 2~hours).

The above results in the papers discussed indicate that (1) the mass density in the at geosynchronous distances is best correlated with $F10.7$ and (2) there exists a much weaker statistical relationship between mass density and the geomagnetic activity indices $K_p$ and $D_{st}$.  In this work we consider the dependence of mass density estimates in the \cite{Denton2006} data set and attempt to identify and statistically characterize the relationship between geomagnetic activity with mass density at geosynchronous distances.  In addition, we attempt to identify solar wind processes that may drive changes in mass density.

%There are several issues that are addressed in detail: (1) the plasma density measurements are sparse, (2) The plasma trough density has a high correlation with F10.7, and geomagnetic and solar wind parameters also correlate with F10.7, (3) the dependence of mass density on the north/south component of the solar wind magnetic field is expected to be nonlinear. For northward IMF, corresponding to low solar wind driving of the magnetosphere, the plasmasphere can expand past geosynchronous orbit where trough estimates of mass density are made resulting in an increase in observed density that are not due to solar wind driving; for southward IMF, events studies have shown an increase in mass density associated with strong southward IMF and enhanced magnetospheric activity (\cite{Takahashi2010}; \cite{Yao2008}).

%Issue (1) is addressed by considering in detail the number of observations that are used in epoch averages around the time of two types of events, decreases in $D_{st}$ below a threshold, and increases in $\rho_{eq}$ above a threshold.  Issue (2) is addressed using various methods to remove or isolate the F10.7 dependence.  Issue (3) is addressed by considering epochs and linear correlations independent of the direction of the IMF along with separation of the epoch averages based on the northward and southward IMF.

%In addition, we consider the influence of time averaging by comparing results using daily means and medians and hourly means and medians of solar wind and magnetospheric parameters.

@@

The parameters $\rho_{eq}$ and $F_{10.7}$ are from the dataset of \cite{Takahashi2010}; data are available from 1980 through 1991 from GOES 2, 3, 5, 6 and 7.  All other parameters used are from \cite{Kondrashov2014ReconstructionOfGaps} which has coverage from 1972 through 2013, which are on a 1--hour time grid. $\rho_{eq}$ is the inferred equatorial mass density based on the 3rd harmonic torodial frequency of magnetic field measurements as described in \cite{Takahashi2010}.  The smallest cadence for $\rho_{eq}$ values is 10 minutes.  To compute an hourly median over the same interval for which the solar wind parameters were averaged, the median of all $\rho_{eq}$ values in a given hour window was used. GOES 6 had the longest span of available data and the primary results presented are for GOES 6 while measurements from other satellites were used for consistency checks.

%Fill (NaN) values were used for hours in which no measurements were available.  In cases where $\rho_{eq}$ was available from multiple GOES satellites at the same time, the value from GOES 6 was selected.  In the dataset, measurements from GOES 6 had the longest time span of coverage.  An overview of the data used is shown in Figure~\ref{AllData}.

Figure~\ref{ccplot} shows the long-term trends of log$_{10}(\rho_{eq})$ from GOES 6 and $F_{10.7}$ computed using the median values in 27-day non-overlapping windows.  A scatter plot of these two lines is shown in the lower panel of Figure~\ref{ccplot} along with that for 1--day and 1--hour medians.  The linear correlation for GOES 6 using all measurements is found to be $0.94$, which is the same value documented by \cite{Takahashi2010} who used measurements from all satellites in the time interval of 1980 through 1991 using the constraint $0600 \leq MLT \le 1200$, $K_{p3d}\ge1.0$ and $D_{st}\ge -50$~nT.  Our 27-day correlations for GOES 2, 5, and 7 are 0.81, 0.78, and 0.87, and the time range of coverage of measurements from these satellites is 1980-01-01/1983-05-16, 1983-01-01/1987-02-28, and 1983-05-27/1991-08-29 respectively.

@@

Two types of events are considered. The first is a drop in the $D_{st}$ index below the threshold of -50~nT. These types of events are considered in order to determine if there is a well-defined $\rho_{eq}$ dependence on geomagnetic activity/storms as indicated by the $D_{st}$ index.  It is expected that the plasmasphere will exhibit two responses near these threshold crossings.  First is the movement of the plasmapause earthward (\cite{LemaireEarthsPlasmasphere}) and the second is a possible change in density due to magnetospheric processes.

@@

Our first analysis uses data only from the time interval 1989-1991, which corresponds with the interval used in \cite{Takahashi2010}. In this interval, there were 329 $D_{st}$ events.  For both types of events, the hour of the threshold crossing defines the zero epoch time, and for each event, $4\cdot24$ hours were considered before and $8\cdot24$ hours after each event.  To compute the epoch averages on a daily time scale, the median value of all available measurements for all events centered on a window of $\pm 12$ hours of the epoch zero hour was computed, and these averaging windows were shifted in increments of $24$~hours. Similar results are obtained if we first reduce each event time series to have a 1-day cadence by computing medians for each event in 1-day bins and then compute the medians across events.

The stack plot in the upper panel of Figure \ref{DailyAveragedDstEvents} shows the epoch averages for the 329 events from 1989-1991.  The minimum $D_{st}$ median is $-48$~nT.  Consistent with \cite{Takahashi2010}, $D_{st}$ events correspond to elevated $\rho_{eq}$ before and after the event, although the magnitude of increase observed here is $1.5$ amu/cm$^3$ (from 18 to 19.5~amu/cm$^3$) instead of the increase of $\sim 10$ amu/cm$^3$ found in Figure 11 of \cite{Takahashi2010}.  The vertical green bars in the $\rho_{eq}$ plot show the number of $\rho_{eq}$ values that were used to compute the medians and red error lines.  Error lines for each parameter are the standard deviation of the values used in computing the median divided by the square root of the number of values. For the non-$\rho_{eq}$\ medians, the number of measurements used in computing the medians is typically equal to the number of events, which is much larger than the number of $\rho_{eq}$\ values as all available measurements were used instead of restricting to only values where a $\rho_{eq}$ value also existed.

In the lower panel of Figure \ref{DailyAveragedDstEvents} it is shown that the trends seen in the analysis of 1989-1991 also hold for the entire span of measurements from GOES 6 - a small elevation on the day of the event relative to the day before and day after, with median averages after the day of the event being slightly lower than that before.

To determine if the magnitude of the density medians in the top panel of Figure~\ref{DailyAveragedDstEvents} on epoch days -1, 0, and 1 are statistically different from each other, we used a two-population bootstrap test.  A bootstrap sample of the medians for each epoch day was created by sampling the values used to compute the median on each epoch day with replacement.  The observed difference between the medians is shown in Table~\ref{BootstrapDifferenceTable} along with the fraction of observations in 1000 bootstrap differences with a different sign than that observed.

If a statistically significant difference in the medians between two days existed, the fraction of bootstrap samples with a different sign is expected to be much smaller ($<$ 5\%).  In terms of a hypothesis test, we cannot reject the null hypothesis that the medians are the same with a confidence higher than 100 minus the \% value shown in the table; the highest confidence level of 87.6\% is in the difference between day 0 and day 1.

@@

\begin{table}
\small
%\input{../tables/DeltaBootstraps-case10.txt}
\caption{Results of test on of means of $\rho_{eq}$ shown in the top panel of Figure~\ref{DailyAveragedDstEvents} between days of threshold crossing  near (day = 1 or -1) or on the day of a $D_{st}$ event (day = 0).}
\label{BootstrapDifferenceTable}
\end{table}

We have also considered hourly averages as shown in the top panel of Figure~\ref{HourlyAveragedDstEvents}.  A total of 668 $D_{st}$ events were found in the interval of May 1983 through August 1991 with an average duration of 9 hours and a median duration of 3 hours.    In this plot, the overall trend is similar to that found in the bottom panel of Figure~\ref{HourlyAveragedDstEvents} - $\rho_{eq}$ is slightly elevated at the time of the threshold crossing of $D_{st}$ with respect to the values one day before and after, but the elevation is small relative to the error bars around the time of the event onset.

The bottom panel of Figure~\ref{HourlyAveragedDstEvents} contains the events shown in the top panel under the constraint that $D_{st}$ remained below -50~nT for at least twelve hours after the event onset time.  In the time period of elevated mass density within 12 hours of the start of the event, there are only $\sim$5 samples per hourly interval.  This result is consistent with previous observations of large mass density spikes in the plasma trough region during large geomagnetic storms (\cite{Yao2008}; \cite{Takahashi2010}).

To determine of this relationship can be also observed in the events of the top panel of Figure~\ref{HourlyAveragedDstEvents}, we sorted the events by the average $D_{st}$ values in the time interval of 0-12 hours, split the events in two and recomputed the epoch averages of $\rho_{eq}$.  The result is that there is no statistically significant difference in the mass density epoch averages for large versus small events.  This indicates that to confirm a statistical $\rho_{eq}$ dependence on large geomagnetic storms would require a dataset with a much larger number of strong geomagnetic storms.

@@

An alternative type of event can be defined that corresponds to large increases in $\rho_{eq}$.  Figure~\ref{HourlyAveragedRhoEvents} shows the epoch time series for events in which an increase of $\rho_{eq}$ above $20$~amu/cm$^2$ defines the start of the event. In finding $\rho_{eq}$ events, missing values were replaced with linearly interpolated values.  These events are associated with positive $B_z$ and low values of $D_{st}$.  Approximately 8 hours after onset, $\rho_{eq}$ peaks and then slowly decays after 40 hours to pre--onset levels.

Figure~\ref{HourlyAveragedRhoEvents} can be compared with Figure~9 of \cite{Denton2016}, which shows daily averages of $\rho_{eq}$ from this dataset after the onset of quiet $K_p$ intervals.  The growth shown here is approximately linear and lasts $\sim$ 12--18 hours whereas the 10 quiet $K_p$ events considered by \cite{Denton2016} showed growth that lasted for at least 48~hours.
